Predictive modeling using genomic annotations

### 2018-09-12 Manuscript/manuscript.Rmd
updates

The types of models and the feature characteristics used in the aforementioned methods vary considerably. Because of the different cell lines available for analysis and the widening list of functional genomic elements, it can be quite useful to have a unified approach in predicting the formation of TAD boundaries. Therefore, in this study we have developed a novel data pre-processing pipeline that addresses the irregularities in the TAD boundary data provided by contact matrices at different resolutions. We then performed a variety of machine learning algorithms on the pre-processes data to identify the genomic elements that were most predictive of TAD boundaries. In our models we incorporated the distances from bin to region as features. We found that the models performed on our pre-processed data performed better than previously established methods. Additionally, we were able to confirm that the cis-regulatory elements like insulators and promotors were highly predictive of TAD boundaries. The specific transcriptional repressor, CCCTC-binding factor (CTCF) was found to be highly predictive of TAD boundaries. Other non-transcription factor sites like DNase I hypersensitive sites and certain histone modifications like H3K36me3 were also found to be better predictors.  Our findings confirm that models with pre-processed data and distance-related genomic features not only perform better, but are also able to correctly identify a range of different functional genomic elements associated with the formation of TAD boundaries. This knowlege will allow us to gain better insight into the role that the spatial organisation of the human genome has on gene regulation.

@@

Publicly available topologically associating domain data was obtained from GEO with accession GSE53525 for the GM12878 and K562 cell lines.  The domain data was constructed from in situ Hi-C contact matrices at a 5kb and 10 kb resolution for the GM12878 and k562 cell lines respectively. The Arrowhead algorithm was used to identify TAD boundaries for a given contact matrix (Rao et. al). Identified TAD boundaries were represented by their genomic coordinates (hg19 human genome assembly), including chromosome, start and end coordinates. The start and end coordinates were concatenated, sorted, and unique coordinates representing unique TAD boundaries were obtained [??? Check for accuracy; !!! what do you mean by accuracy here?]. The establishment of the data for modeling first consisted of binning the genome into equal sized intervals. This was performed at the same resolution as the respective cell line. That is, for GM12878, 5 kb genomic bins were used, while 10 kb intervals were used for K562. Next, for the domain data provided by each cell line, the boundary coordinates were flanked on either side by 5 kb [!!! use the same size flank for k562? (resolution was 10kb)]. A genomic bin was labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing an boundary (Y=0).[??? Should it be simplified - just consider boundaries to be flanked?; !!!].

@@

Annotation data was obtained from the Encyclopedia of DNA Elements (ENCODE) Consortium [??? We need a table. Check Hong 2017 supplemental, there may be something else that I don't have in GenomeRunner database. Also, check iGD data http://big.databio.org/igd/; !!! table in supplementary file, Hong considers TFBS, should we also include?].  These annotations consisted of genomic coordinates and were used to build the feature space, X = {$X_{1}$, . . ., $X_{p}$}, of the subsequent models. The full list of features is provided in [X]. For each genomic feature, two predictors types were considered, a count variable and a distance variable. The count variable was based on the number of regions of each functional element that overlapped with a particular genomic bin. The distance variable consisted of the length (in base pairs) from the center of each genomic bin to the center of the nearest region of a functional element. If the center of the region given by a particular functional element was located within a genomic bin, then the distance was given to be 0.[??? Measured in?; !!!]. Figure X presents an illustration for how both the response vector and feature space were constructed.

@@

We developed a pipeline that systematically evaluates the problems associated with genomic data in predictive modeling settings.

At each stage in the pipeline, Elastic-Net models were performed and performance metrics such as ROC curves and subsequent AUCs were evaluated. The Elastic-Net model was used here because of its ability to be tuned and its computational speed relative to other machine learning algorithms. For each elastic-net model the mixing parameter, alpha, was tuned on a grid of five values (0.1, 0.325, 0.550, 0.775, and 1.0). A sequence of values for the regularization parameter, lambda, was automatically provided by the CARET package in R. The algorithm then chose the optimal combination of alpha and lambda, and the model was evaluated. [??? Unclear, is it cross-validation tuning or what?]

Prior to initiating the pipeline, predictors with near-zero variance were removed. To identify said predictors, the nearZeroVar function in the CARET package was used. The function considers two metrics. First, the ratio of the most frequent value over the second most frequent value was calculated. Ideally, this ratio would be close to one for well-behaved predictors. Next, the number of unique values divided by the total number of samples multiplied by 100 was calculated. Here, if the frequency ratio was greater than 20 and the percentage of unique values was less than 10, the predictor was flagged as having near-zero variance and removed.

@@

To assess the class imbalance problem due to the sparcity of domains throughout the genome, two methods were used. The first consisted of incorporating the SMOTE command from the DmWR package in R. SMOTE stands for Synthetic Minority Over-sampling Technique. The SMOTE algorithm incorporates both under-sampling (percent under) of the majority class and over-sampling (percent over) of the minority class. The minority class is over-sampled by taking each minority class sample and introducing synthetic examples along the line segments joining any/all of the k minority class nearest neighbors (here, k=5). Then, the majority class is under sampled by taking a random sample that matches some pre-specified percentage of the updated minority class.  For example, consider a data set with 700 observations, where there are 35 minority classes and 665 majority classes. Now, when using SMOTE, consider a 200 percent over sample of the minority class and a 100 percent under sample of the majority class.  This means that two synthetic observations will be created for every minority class (35+70=105) and 1 majority class will be randomly chosen for every new minority class created (70), resulting in a classification of 70/105 (majority/minority). Now consider, a 200/200 combination. The new minority class is still 105 (70 new synthetic observations created), but now twice as many majority classes are randomly chosen (70*2=140), resulting in a 140/105 classification. It is important to note that using a combination of 100/200 will always result in perfectly balanced classes. A total of eight different combinations of percent over and undersampling were used which included: 100/200, 200/200, 300/200, 400/200, 100/300, 200/300, 300/300, 400/300. Here, the 100/200 combination creates a completely balanced dataset as mentioned before, while the others create an unbalanced dataset with the “No” class as the majority to reflect the original data. The second method involved iteratively taking 100 bootstrap samples of the majority class to match the number in the minority class. For each sample, an elastic-net model was tuned and evaluated. Then row wise means of the sensitivities, specificities, and AUCs of each of the 100 bootstrap samples were taken to evaluate the models as a whole. The SMOTE combination with the best performance was then compared to the bootstrap method. ROC curves were plotted and corresponding AUCs were presented in a table.

In order to correct for the skewness of the distances [??? Should be defined in the introduction], elastic-net models were evaluated based on a combination of log transformed and/or standardized predictors. A total of four elastic net models were performed including: with log transformation and standardization, without log transformation and with standardization, with log transformation and without standardization, and without log transformation and without standardization. One hundred bootstrap samples were taken from the majority class to create a balanced classification data. ROC curves were plotted and corresponding AUCs were presented in a table.  Likewise, we looked at the variable importance quality of the models.

@@

Due to the large number of genomic features included in the model, different variable selection techniques were evaluated in order to prevent overfitting, improve computational speed, and eliminate uninformative variables. There were three different wrapper methods that were used here, forward, backward, and stepwise selection. A null model consisting of just an intercept term and a full model, including all features, were specified for each selection technique. For all three, a logistic regression model was performed, and the AIC of the fitted model was used to establish which variables to include. That is, only a reduction in the model's AIC contributed to the inclusion of a feature in the model. Additionally, a recursive feature elimination (RFE) algorithm was performed. In RFE, a full model was created, a measure of variable importance was then computed that ranked the predictors from the most important to the least. Here, the importance of the model was based on the random forest importance criterion [??? Unclear how is it different from backward selection, compare and contrast]. Then, at each stage of the search, the least important predictors were iteratively eliminated prior to rebuilding the model. The process was continued for a pre-specified subset of predictors.

>>>>

The types of models and the feature characteristics used in the aforementioned methods vary considerably. Because of the different cell lines available for analysis and the widening list of functional genomic elements, it can be quite useful to have a unified approach in predicting the formation of TAD boundaries. Therefore, in this study we have developed a novel data pre-processing pipeline that addresses the irregularities in the TAD boundary data provided by contact matrices at different resolutions for two cell lines, GM12878 and K562. We then performed a variety of machine learning algorithms on the pre-processes data to identify the genomic elements that were most predictive of TAD boundaries. In our models we incorporated the distances from bin to region as features. We found that the models performed on our pre-processed data performed better than previously established methods. Additionally, we were able to confirm that the cis-regulatory elements like insulators and promotors were highly predictive of TAD boundaries. The specific transcriptional repressor, CCCTC-binding factor (CTCF) was found to be highly predictive of TAD boundaries. Other non-transcription factor sites like DNase I hypersensitive sites and certain histone modifications like H3K36me3 were also found to be better predictors.  Our findings confirm that models with pre-processed data and distance-related genomic features not only perform better, but are also able to correctly identify a range of different functional genomic elements associated with the formation of TAD boundaries. This knowlege will allow us to gain better insight into the role that the spatial organisation of the human genome has on gene regulation.

@@

Publicly available topologically associating domain data was obtained from GEO with accession GSE53525 for the GM12878 and K562 cell lines.  The domain data was constructed from in situ Hi-C contact matrices at a 5kb and 10 kb resolution for the GM12878 and k562 cell lines respectively. The Arrowhead algorithm was used to identify TAD boundaries for a given contact matrix (Rao et. al). Identified TAD boundaries were represented by their genomic coordinates (hg19 human genome assembly), including chromosome, start and end coordinates. The start and end coordinates were concatenated, sorted, and unique coordinates representing unique TAD boundaries were obtained [??? Check for accuracy; !!! what do you mean by accuracy here?]. A depiction of the data is presented in Figure X.

The establishment of the data for modeling first consisted of binning the genome into equal sized intervals. This was performed at the same resolution as the respective cell line. That is, for GM12878, 5 kb genomic bins were used, while 10 kb bins were used for K562. Next, for the domain data provided by GM12878, the boundary coordinates were flanked on either side by 5 kb. For the K562, the boundary coordinates were flanked by 10 kb. In either case, a genomic bin was labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing an boundary (Y=0) (Figure X).

@@

Annotation data was obtained from the Encyclopedia of DNA Elements (ENCODE) Consortium.  These annotations consisted of genomic coordinates and were used to build the feature space, X = {$X_{1}$, . . ., $X_{p}$}, of the subsequent models. A full list of the genomic elements is provided in Table X. For each genomic element, the distance from the center of a genomic bin to the center of the nearest region (in base pairs) was calculated and included in the model.  If the center of the region given by a particular functional element was located within a genomic bin, then the distance was given to be 0. Additionally, a variable denoting the number of overlaps between a genomic bin and region representing a genomic element was totaled and included in the model (Figure X).

@@

We developed a pipeline that systematically evaluates the problems associated with TAD boundary data in a predictive modeling setting (Figure X). The pipeline was used to address issues related to normalization of predictors, class imbalance, and variable reduction. Prior to initiating the pipeline, predictors with near-zero variance were removed. To identify said predictors, the nearZeroVar function in the CARET package was used. The function considers two metrics. First, the ratio of the most frequent value over the second most frequent value was calculated. Ideally, this ratio would be close to one for well-behaved predictors. Next, the number of unique values divided by the total number of samples multiplied by 100 was calculated. Here, if the frequency ratio was greater than 20 and the percentage of unique values was less than 10, the predictor was flagged as having near-zero variance and removed.

At each subsequent stage in the pipeline, Elastic-Net models were performed and performance metrics such as F1 scores and MCCs were evaluated. The Elastic-Net model was used here because of its ability to be tuned and its computational speed relative to other machine learning algorithms. To avoid overfitting and more accurately assess performance, 10-fold cross validation was used for all models. For each elastic-net model the mixing parameter, alpha, was tuned over a grid of 10 evenly spaced values from 0 and 1. The penalty term, labmda, was tuned over a grid of 100 evenly spaced values from 0.01 to 100.

Distances on a genomic scale are frequently highly skewed. In order to correct for this, a log base 2 transformation was considered. Additionally, the predictors were standardized to have unit variance in order to be able to directly compare magnitudes of parameter coefficients.

@@

When binning the genome, the sparcity of TAD boundaries through the genome contributed to heavily imbalanced classes. That is, proportionally, only a small number of genomic bins overlapped with a flanked TAD boundary. To assess this issue, two methods were used. The first consisted of incorporating the SMOTE command from the DmWR package in R. SMOTE stands for Synthetic Minority Over-sampling Technique. The SMOTE algorithm incorporates both under-sampling (percent under) of the majority class and over-sampling (percent over) of the minority class. The minority class is over-sampled by taking each minority class sample and introducing synthetic examples along the line segments joining any/all of the k minority class nearest neighbors (here, k=5). Then, the majority class is under sampled by taking a random sample that matches some pre-specified percentage of the updated minority class. It is important to note that using a combination of 100/200 (percent over/percent under-sample) will always result in perfectly balanced classes. A total of four different combinations of percent over and under-sampling were used which included: 100/200, 200/200, 300/200, and 400/200. As mentioned before, the 100/200 combination created balanced classes, while the others create an unbalanced dataset with the negative class as the majority to reflect the proportionality of the classes in the original data. The second method involved iteratively under-sampling 100 time from the majority class to match the number in the minority class. For each sample, an elastic-net model was tuned and evaluated. Then row-wise means of the performance metrics for each of the 100 interations were taken to evaluate the models (Figure X).

@@

Due to the large number of genomic features included in the model, different variable selection techniques were evaluated in order to prevent overfitting, improve computational speed, and eliminate uninformative variables. There were three different wrapper methods that were used here, forward, backward, and stepwise selection. A null model consisting of just an intercept term and a full model, including all features, were specified for each selection technique. For all three, a logistic regression model was performed, and the AIC of the fitted model was used to establish which variables to include. That is, only a reduction in the model's AIC contributed to the inclusion of a feature in the model. For each technique, prior to model selection, balanced classes were iteratively created 100 times by random under-sampling from teh majority class. At each iteration the variables selected from each technique were recorded. The final set of predictors selected for each technique were those that appeared over 90% of the time accross all iterations. The union of the three sets of features from each of the techniques was obtained and used as the final feature space of the data (Figure X)
